---
title: "Gapminder"
author: "Ashish Yadav"
output:
  html_notebook:
    output:
      toc: yes
      toc_float: null
  highlight: breezedark
  collapsed: no
  html_document:
    df_print: paged
  theme: cosmo
---
The objective of this case study is to understand the change in World Health and Economics using Data Visualization, EDA, and Summarization. In this study two main questions, *Is it a fair characterization of today's world to say that it is divided into a Westorn Rich Nations (Europian Countries, USA et cetera), and Developing Countries (Asia, Africa et cetera)? Has the **Income Inequality** worsened during the last 40 years?* The study involves data from Gapminder Foundation about trends in world health and economics. Study emphasizes the use of data visualization to better understand the trends and insights. This study is purely based on the Gapminder TED talks **New Insights on poverty**

## Load and explore data

```{r}
library(dslabs)
library(tidyverse)
data(gapminder)
gapminder %>% head()
str(gapminder)
```

**Data consists of 10545 Observations and 9 Variables**, it consists of varibles like *country*, *region*, health outcomes (*life_expectancy*, *fertility*), economic aspects (*gdp*), 
**Add gdp_cp var i.e. gdp per capita which represents the wealth of a country**

```{r}
gapminder <- gapminder %>% mutate(gdp_pc = gdp / population, dollars_per_day = gdp_pc/365)
head(gapminder)
```

## Analysis
### Infant Mortality
Getting started with testing our knowledge regarding differences in infant mortality across differnt countries, for each of the pairs of countries given below, **Which country do you think had the highest child mortality rate in 2015?** and **Which pairs do you think are the most similar?**

|Country1 |Country2|
|---------|--------|
|Sri Lanka|Turkey  |
|Poland|South Korea|
|Malaysia|Russia|
|Pakistan|Vietnam|
|Thialand|South Africa|

It is commonly percieved that the non-europian countries like Sri Lanka, South Korea have higher mortality rates than their Europian counterparts. Also the developing countries like Pakistan are considered to have high mortality rates. Lets take a look at the data to see whether it is just a superstition or a fact.
```{r}
countries <- c("Sri Lanka","Turkey","Poland","South Korea","Malaysia","Russia","Pakistan","Vietnam","Thailand","South Africa")
mortality <- data.frame()
for (i in seq(1,10,2)) {
    mortality1 <- gapminder %>%
        filter(year == 2015 & country %in% countries[c(i,i+1)]) %>%
        select(country, infant_mortality)
    mortality1 <- cbind(mortality1[1,], mortality1[2,])
    mortality1 <- mortality1[-2,]
    mortality <- rbind.data.frame(mortality,mortality1)
}
names(mortality) <- c("Country1", "Mortality1", "Country2", "Mortality2")
mortality
```
We see that the European countries on this list have higher child mortality rates: Poland
has a higher rate than South Korea, and Russia has a higher rate than Malaysia. We also
see that Pakistan has a much higher rate than Vietnam, and South Africa has a much
higher rate than Thailand. The reason for this stems from the preconceived notion that the world is divided into two
groups: the western world (Western Europe and North America), characterized by long
life spans and small families, versus the developing world (Africa, Asia, and Latin America)
characterized by short life spans and large families.

### Life Expectancy, Fertility
scatterplot of life expectancy versus
fertility rates (average number of children per woman) 50 years ago
```{r}
ds_theme_set()
gapminder %>%
    filter(year == 1962) %>%
    ggplot(aes(fertility, life_expectancy)) +
    geom_point(size = 2) +
    ylab("Life Expectancy") +
    xlab("Fertility")
```

Most points fall into two distinct categories: Life expectancy around 70 years and 3 or fewer children per family, and Life expectancy lower than 65 years and more than 5 children per family.

To confirm that indeed these countries are from the regions we expect, we can use color to
represent continent.
```{r}
gapminder %>%
    filter(year == 1962) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point(size = 2) +
    ylab("Life Expectancy") +
    xlab("Fertility")
```
In 1962, “the West versus developing world” view was grounded in some reality. Is this still
the case 50 years later?

### Changes over time
Facet life expectancy vs fertility by continent and year to see how it changed from 1962 to 2012 for different continents using side-by-side plots
```{r}
gapminder %>% filter(year %in% c("1962", "2012")) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point(size = 2) +
    facet_grid(continent ~ year) +
    ylab("Life Expectancy") +
    xlab("Fertility")
```

Except for countries in Africa continent almost all of the countries had significant increase in life expectancy and reduced fertility and Europian countries has the most significant increase of all thus **It's quite clear from the plot, notion that Europian and American countries have a higher life-expectancy is somewhat correct

### Facet by year only
```{r}
gapminder %>% filter(year %in% c("1962", "2012")) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point(size = 2) +
    facet_grid(.~ year) +
    ylab("Life Expectancy") +
    xlab("Fertility")
```

Faceting by year 1962 and 2012 shows, though all of the countries had a increase in life-expectancy, European and American countries had the highest life-expectancy.
This plot clearly shows that the majority of countries have moved from the developing world
cluster to the western world one. In 2012, the western versus developing world view no
longer makes sense. This is particularly clear when comparing Europe to Asia, the latter of
which includes several countries that have made great improvements.


**Facet by year, plots wrapped onto multiple rows to see changes over the years in life-expectancy** to explore how this transformation happened through the years, we can make the plot for
several years. This plot clearly shows how most Asian countries have improved at a much faster rate than
European ones.

```{r}
y <- c("1962", "1980", "1990", "2000", "2012")
c <- c("Europe", "Asia")
gapminder %>% filter(year %in% y & continent %in% c) %>%
    ggplot(aes(fertility, life_expectancy, col = continent)) +
    geom_point(size = 1) +
    facet_wrap(~year) +
    ylab("Life Expectancy") +
    xlab("Fertility")
```

## Time Series Analysis
The visualizations above effectively illustrate that data no longer supports the western versus
developing world view. Once we see these plots, new questions emerge. For example, which
countries are improving more and which ones less? Was the improvement constant during
the last 50 years or was it more accelerated during certain periods? For a closer look that
may help answer these questions, we are going to use time series plots.

```{r}
gapminder %>%
    filter(country == "United States", !is.na(fertility)) %>%
    ggplot(aes(year, fertility)) +
    geom_point()
```
We see that the trend is not linear at all. Instead there is sharp drop during the 1960s and
1970s to below 2. Then the trend comes back to 2 and stabilizes during the 1990s.
When the points are regularly and densely spaced, as they are here, we create curves by
joining the points with lines, to convey that these data are from a single series, here a
country.

### Line plot - US fertility
```{r}
gapminder %>%
    filter(country == "United States", !is.na(fertility)) %>%
    ggplot(aes(year, fertility)) +
    geom_line()
```

### Line plot - Korea and Germany
This is particularly helpful when we look at two countries. If we subset the data to include
two countries, one from Europe and one from Asia.

```{r}
countries <- c("South Korea", "Germany")
labels <- data.frame(country = countries, x = c(1986, 1975), y = c(2.5,2.0))
gapminder %>% filter(country %in% countries & !is.na(fertility)) %>%
    ggplot(aes(year, fertility, col = country)) +
    geom_text(data = labels, aes(x, y, label = country), size = 4) +
    theme(legend.position = "none") +
    geom_line()
```
The plot clearly shows how South Korea’s fertility rate dropped drastically during the 1960s
and 1970s, and by 1990 had a similar rate to that of Germany.

```{r}
labels <- data.frame(country = countries, x = c(1975,1965), y = c(60,72))
gapminder %>%
filter(country %in% countries) %>%
ggplot(aes(year, life_expectancy, col = country)) +
geom_line() +
geom_text(data = labels, aes(x, y, label = country), size = 5) +
theme(legend.position = "none")
```
The plot clearly shows how an improvement in life expectancy followed the drops in fertility
rates. In 1960, Germans lived 15 years longer than South Koreans, although by 2010 the
gap is completely closed. It exemplifies the improvement that many non-western countries
have achieved in the last 40 years.

